---
title: "Final Biostatistics Analysis"
output: html_notebook
---

# Biostatistics Analysis 2: Australian Fires

## Step 1: Loading the tidyverse

```{r}
#Load the tidyverse
library(tidyverse)
library(kableExtra)
#install.packages("tidymodels")
library(tidymodels)
#install.packages("skimr")
library(skimr)
```

## Step 2: Loading the packages

The data set I chose for this analysis is on Australia, specifically about it's climate and recent fires. One of the goals of compiling this data is to spread awareness of these fires, and there is an

```{r}
# Get the Data

rainfall <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-07/rainfall.csv')
temperature <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-01-07/temperature.csv')

library(skimr)
rainfall%>%
  skim()

library(skimr)
temperature%>%
  skim()

rainfall %>%
  head() %>%
  kable() %>%
  kable_styling(c("hover", "striped"))
```

Above I have loaded the data from the tidytuesday page, which has two data sets (temperature and rainfall).
