---
title: "VolcanicEruptionsAnalysis"
author: "Melanie Dube"
format: html
editor: visual
execute: 
  keep-md: true
---

# Volcanic Eruptions Analysis

## Abstract

This notebook examines data from a number of volcanic eruptions over the course of 2,500 years. There are a number of questions which can be generated from this data, including which volcano has the most eruptions, what region of the world has the most active volcanoes, what was the longest confirmed eruption, etc..

## Introduction

There are 5,589 total volcanoes which were have been added to this data set thus far. The observed volcanoes are located across the world, spanning all seven continents. There are a variety of classifications which volcanoes can fall into, including active, dormant, and extinct. The exploratory function for this set allows access to many different categories of data. The name and identification numbers (including eruption number and volcano number) for each volcano are listed, as well as eruption category. Additional data about the start and end dates of each eruption, and physical location of the volcano are also tracked in the data table.

### Reading the Data into Notebook

Always begin a new set of r with the library(tidyverse) code, as well and the data packages, which are pasted below.

Begin each new session by running the tidyverse and the package of interest.

```{r}
library(tidyverse)
#install.packages("skimr")
library(skimr)

volcano <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')

```

```{r}
#install.packages("tidymodels")
library(tidymodels)

my_data_splits <- initial_split(eruptions, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
```

We have specified eruptions (in the initial_split function), so that is the data which will be available. By changing this to one of the other five available data sets in the initial load, we can view different data.

By running exploratory_data, the table below is generated.

```{r}
exploratory_data
```

### Interesting Questions to Ask

-   What is the name of the volcano with the largest number of eruptions?

-   What percent of the total eruptions were confirmed vs. unconfirmed?

-   What are all different evidence methods and how many times were each of them used?

-   Are volcanic eruptions more likely to happen in one area of the globe?

## Data Exploration

```{r}
exploratory_data %>%
  skim(evidence_method_dating, data_name = NULL)
```

This uses the skim function to summarize data, by skimming the "evidence_method_dating" column we can see that there are 5589 rows, which should correspond to 5589 individual tracked eruptions. There are 15 columns, which could correspond to 15 different methods of evidence dating. This function does not show each of the names of the data collection methods.

```{r}
exploratory_data %>%
  count(eruption_category)
```

This function shows us how many of each eruption category are tracked in this data set. There appear to be 4,964 confirmed eruptions, 80 discredited eruptions, and 545 uncertain eruptions. This could be useful data because in determining answers to questions about actual eruptions, it may be necessary to only look at the 4,964 eruptions which were confirmed.

```{r}
exploratory_data %>%
  count(volcano_number)

exploratory_data %>%
  count(volcano_number) %>%
  arrange(-n)
```

This counts the eruptions per individual volcano. By using the arrange function, we are able to change the way that the data is presented and organize it in descending order, so that the volcano with the most eruptions is listed first. This shows us that volcano number 211060 has erupted the most times (124). Because this is so much more than any of the others, it can be considered an out liar and could be worth investigating more.

```{r}
exploratory_data %>%
  group_by(volcano_name) %>%
  count(eruption_category) %>%
  arrange(-n)


```

By running the code above we are able to see which volcanoes have the highest number of confirmed eruptions. It is evident that Etna has 100 confirmed eruptions, but there are at least 10 other volcanoes which have more than 50 confirmed eruptions.

```{r}
#install.packages("leaflet")
#install.packages("ggmap")

library(leaflet)
confirmed <- exploratory_data %>%
  filter(eruption_category == "Confirmed Eruption")
leaflet() %>%
  addProviderTiles(providers$OpenTopoMap) %>%
  addCircleMarkers(lng = confirmed$longitude, lat = confirmed$latitude, radius = 0.5)
```

```{r}
top10erupted <- exploratory_data %>%
  group_by(volcano_name, latitude, longitude) %>%
  count(eruption_category) %>%
  arrange(-n) %>%
  head(10)
leaflet() %>%
  addProviderTiles(providers$OpenTopoMap) %>%
  addCircleMarkers(lng = top10erupted$longitude, lat = top10erupted$latitude, radius = top10erupted$n/10)
```

### Hypotheses

-   I hypothesize that volcanoes with the highest number of confirmed eruptions do not all exist in one region of the world.

-   I hypothesize that the most common method of evidence dating is Historical Observations.

### Proving Hypotheses

-   The first hypothesis can be proved with the code block above. Using ggplot, we are able to place the ten volcanoes with the highest recorded amounts of confirmed eruptions on a topographic map. We are able to see that the volcanoes are located across the map, not just in one region. Of the top ten, there are none existing in North America. There appear to be three of the top ten located in the far eastern Asia/Japan region, and another in the Indonesia. Based on this data we may be able to make inferences about where some of the most active volcanoes in the world are located.

-   The second hypothesis is examined in the code block below. It is evident that of the 20 different methods of evidence dating, historical observation is the most commonly used. However, it is worth noting that Radiocarbon and Tephrochronology were used a significant number of times as well, accounting for a total of 1436 out of 4964 total confirmed eruptions, or 29%. Additionally, 626 out of 496 confirmed eruptions (13%) have an unknown method, which could also be noted for significance.

```{r}
exploratory_data %>%
  count(evidence_method_dating) %>%
  arrange(-n)
```

## Inference/Methodology

In the inference section of this data analysis, a new set of data is used. Instead of using exploratory data, the below section shows the test data, which is completely separate from any data used previously.

```{r}
test_data
```

#### Combining Columns

```{r}
test_data<-test_data %>%
  mutate(east_west = ifelse(longitude>=0, "east", "west")) %>%
  mutate(north_south = ifelse(latitude>=0, "north", "south"))


test_data<-test_data %>%
  unite(col=region, c('north_south','east_west'), sep = "-" )
```

The block above takes data for latitude and longitude and combines them to determine the region of the world where the volcano is located. The mutate function helps determine whether the volcano is north/south or east/west, and the unite function takes created columns for these two variables and combines them into one called "region", where each volcano can be categorized into one of four different regions, north-east, north-west, south-east, or south-west.

#### Eruptions per Region

```{r}
library(ggplot2)

test_data%>%
  ggplot(aes(x=region, y=volcano_number)) +
  geom_bar(stat="identity")
```

The plot above shows each of the regions of the world and the corresponding volcanoes which have been studied in each region. Using this plot we are able to see that the south-west appears to be the least "active" region in terms of volcanic eruptions, while the north eastern region is the most active. In general, there appear to be significantly more eruptions in the north compared to the south.

#### Methods of Evidence Dating

```{r}
dating_counts <- test_data %>% count(evidence_method_dating) %>% pull(n)
dating_counts <- as.numeric(dating_counts)
dating_probs <- rep(1/20, 20)
chisq <- chisq.test(dating_counts, p = dating_probs)

chisq
```

This testing method allows to determine the likelihood of one dating method to be used compared to all others. The null hypothesis we have created is that all evidence method dating types are equally likely. Therefore, the alternative hypothesis would be that all the evidence method dating types are not equally likely. Through a chi-square goodness of fit test, we can compare the observed distribution to an expected distribution. Running this test with the data calculated an extremely small p value (2.2e-16), which leads us to believe that the alternative hypothesis is correct and not all methods are equally likely to be used.

```{r}
test_data %>%
  ggplot(aes(x = fct_reorder(evidence_method_dating,volcano_number), y = volcano_number)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(x = "")
  
                

```

This chart gives a visual for the test performed above, which confirms the alternative hypothesis. Based on the size of the bars, not all methods of testing have equal chances of being used in this test data set.
